---
title: "Data Generation : Teriflunomide case study (time-to-event endpoint)" 
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: cerulean
    toc: yes
    toc_float:
      collapsed: true
vignette: >
  %\VignetteIndexEntry{Data Generation : Teriflunomide case study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{BExTE, ggplot2, survival}
---


```{r setup, warning=F, message=F, include=FALSE} 
root <- dirname(dirname(getwd()))
knitr::opts_knit$set(root.dir = root, warning = FALSE, message = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(BExTE)
library(ggplot2)
```
 

# Teriflunomide case study

Sources: 
[Bovis et al (2022)](https://jamanetwork.com/journals/jamaneurology/fullarticle/2793703)

The source study is the pooled analysis of the adult TEMSO and TOWER trials of
14 mg teriflunomide; Table 4 of the TOWER paper by
[Confavreux et al (2014)](https://linkinghub.elsevier.com/retrieve/pii/S1474442213703089)
reports it. The target study is a paediatric trial inspired by the TERIKIDS
design. The endpoint is the time to first relapse.

The source treatment effect is the pooled log hazard ratio
$\widehat{\theta}_S = \log(\widehat{HR}_S) = -0.393$, with
$\widehat{\sigma}_{\theta_S} = 0.081$. The target treatment effect is
$\theta_T = \widehat{\theta}_S + \delta$, where $\delta$ is the drift.

```{r}
set.seed(42)

config_path <- system.file("conf/case_studies/teriflunomide.yml", package = "BExTE")
case_study_config <- yaml::yaml.load_file(config_path)
format_case_study_config(case_study_config)
```


# The target trial design

In the original TERIKIDS design every participant could be followed for up to
96 weeks, so accrual would not affect the follow-up distribution. To study the
effect of staggered accrual we use a modified fixed-calendar design: the maximum
individual follow-up stays at $L = 96$ weeks, but the database is closed
$F = 48$ weeks after the end of recruitment.

Patients enter over an accrual period of $A = 3.41$ years, with calendar entry
times $R_i \sim \mathcal{U}(0, A)$. The database closes at $C = A + F$, so the
administrative censoring time of patient $i$ is

$$C_i = \min\{L,\ A + F - R_i\}.$$

Control-arm heterogeneity is $\kappa = \log \lambda_T^{(c)} - \log \lambda_S^{(c)}$,
so that $\lambda_T^{(c)} = \lambda_S^{(c)} e^{\kappa}$ and
$\lambda_T^{(t)} = \lambda_T^{(c)} e^{\theta_T}$. In the package $\kappa$ is the
`control_drift` of a scenario and $\delta$ is its `drift`.

Independent loss to follow-up is $D_i \sim \operatorname{Exp}(\lambda_D)$ with
$\lambda_D = -\log(1 - p_D)/L$ for a prespecified probability $p_D$ of being
lost over $L$. The observed follow-up time and the event indicator are

$$X_i = \min(T_i, D_i, C_i), \qquad \Delta_i = \mathbb{1}\{T_i \leq D_i,\ T_i \leq C_i\}.$$

```{r}
design <- case_study_config$target
L <- design$max_follow_up_time
A <- design$accrual_period
Fu <- design$final_follow_up

# Only patients recruited before A + F - L are followed for the full L.
c(max_follow_up = L, accrual = A, final_follow_up = Fu,
  proportion_with_full_follow_up = (A + Fu - L) / A)
```


# Creating the target data

```{r}
source_data <- SourceData$new(case_study_config)

drift <- 0.2 # delta
control_drift <- 0 # kappa
treatment_drift <- drift + control_drift

target_sample_size_per_arm <- 185
```

```{r} 
target_data <- TargetDataFactory$new()
target_data <- target_data$create(
  source_data = source_data,
  case_study_config = case_study_config,
  target_sample_size_per_arm = target_sample_size_per_arm,
  control_drift = control_drift,
  treatment_drift = treatment_drift,
  summary_measure_likelihood = source_data$summary_measure_likelihood,
  dropout_probability = 0.05,
  event_time_distribution = "exponential"
)

c(theta_T = target_data$treatment_effect,
  control_rate = target_data$control_rate,
  treatment_rate = target_data$treatment_rate)
```

The arm rates follow from $\kappa$ and $\theta_T$:

```{r}
read_function_code(rate_from_drift_logRR)
```


# Aggregate data generation

## Generation of aggregate data without sampling approximation

### Usage

```{r}
n_replicates <- 2000
data <- target_data$generate(n_replicates = n_replicates)
head(data)
```

```{r}
target_data$plot_sample(data)
```

### Implementation details

Under the primary proportional-hazards model the time to first relapse is
exponential, $T_i \mid Z_i = a \sim \operatorname{Exp}(\lambda_T^{(a)})$. Each
simulated trial is analysed with a Cox proportional-hazards model
$h_i(t) = h_0(t)\exp\{\beta \mathbb{1}(Z_i = t)\}$, and
$\widehat{\theta}_T = \widehat{\beta}$ with
$\widehat{\sigma}_{\theta_T} = \operatorname{SE}(\widehat{\beta})$ are passed to
the Bayesian borrowing procedures as in the other case studies.

Fitting one `survival::coxph()` per replicate would be far too slow for the
simulation grid, so the package solves the two-sample Cox model directly. With a
single binary covariate the risk set at an event time is described entirely by
the numbers still at risk in each arm, $(n_0, n_1)$, so the score and observed
information reduce to

$$U(\beta) = d_1 - \sum_{\text{events}} \frac{n_1 e^{\beta}}{n_0 + n_1 e^{\beta}},
\qquad
I(\beta) = \sum_{\text{events}} \frac{n_0 n_1 e^{\beta}}{(n_0 + n_1 e^{\beta})^2},$$

which a few Newton steps solve for every replicate at once. Event times are
continuous, so ties have probability zero and this agrees with `coxph()` to
machine precision. The patient-level version below is what the vectorised code
computes:

```{r reference_implementation}
simulate_one_trial <- function(n_per_arm, control_rate, treatment_rate,
                               L, A, Fu, dropout_rate) {
  n <- 2 * n_per_arm
  arm <- rep(c(0, 1), each = n_per_arm)
  rate <- ifelse(arm == 0, control_rate, treatment_rate)

  entry <- runif(n, 0, A)
  administrative <- pmin(L, A + Fu - entry)
  event_time <- rexp(n, rate = rate)
  dropout <- if (dropout_rate > 0) rexp(n, rate = dropout_rate) else Inf

  data.frame(
    time = pmin(event_time, dropout, administrative),
    status = as.integer(event_time <= dropout & event_time <= administrative),
    arm = arm
  )
}

dropout_rate <- -log1p(-0.05) / L

set.seed(123)
trial <- simulate_one_trial(
  n_per_arm = target_sample_size_per_arm,
  control_rate = target_data$control_rate,
  treatment_rate = target_data$treatment_rate,
  L = L, A = A, Fu = Fu, dropout_rate = dropout_rate
)

fit <- survival::coxph(survival::Surv(time, status) ~ arm, data = trial)
c(estimate = unname(coef(fit)), standard_error = unname(sqrt(diag(vcov(fit)))))
```

The package reproduces that fit exactly:

```{r}
vectorised <- BExTE:::fit_two_sample_cox(
  time = matrix(trial$time, nrow = 1),
  event = matrix(trial$status, nrow = 1),
  arm = trial$arm
)
c(estimate = vectorised$estimate, standard_error = vectorised$standard_error)
```

Averaged over replicates, the estimates recover $\theta_T$ and the reported
standard errors match their spread:

```{r}
c(theta_T = target_data$treatment_effect,
  mean_estimate = mean(data$treatment_effect_estimate, na.rm = TRUE),
  mean_standard_error = mean(data$treatment_effect_standard_error, na.rm = TRUE),
  observed_spread = sd(data$treatment_effect_estimate, na.rm = TRUE))
```


## Weibull sensitivity analysis

As a sensitivity analysis, event times are generated from a Weibull
proportional-hazards model with conditional survival function
$S(t \mid Z_i) = \exp\{-(t/b_c)^q \exp(\theta_T Z_i)\}$, where $q$ is the common
shape parameter and $b_c$ the control-arm scale. This implies a hazard ratio of
$\exp(\theta_T)$. We use $q = 0.72$, based on the aggregate TERIKIDS
relapse-free probabilities. The control scale is calibrated to reproduce the
reported 96-week relapse-free probability $S_c(L) = 0.442$, giving

$$b_c = L / [-\log\{S_c(L)\}]^{1/q} = 2.45 \text{ years},$$

and the treatment-arm scale is $b_t = b_c \exp(-\theta_T/q)$. The exponential
model is recovered when $q = 1$.

Because $\exp(\kappa)$ multiplies the control hazard, and multiplying a Weibull
hazard by a constant divides its scale by that constant raised to $1/q$, control
heterogeneity enters as $b_c(\kappa) = b_c \exp(-\kappa/q)$.

```{r}
weibull_data <- TargetDataFactory$new()$create(
  source_data = source_data,
  case_study_config = case_study_config,
  target_sample_size_per_arm = target_sample_size_per_arm,
  control_drift = control_drift,
  treatment_drift = treatment_drift,
  summary_measure_likelihood = source_data$summary_measure_likelihood,
  dropout_probability = 0.05,
  event_time_distribution = "weibull"
)

c(shape = weibull_data$weibull_shape,
  control_scale = weibull_data$weibull_scale,
  relapse_free_at_L = 1 - pweibull(L, weibull_data$weibull_shape,
                                   weibull_data$weibull_scale))
```

Note that the two data-generating models are deliberately not nested: the
exponential arm is calibrated on the source placebo relapse rate, the Weibull
arm on the aggregate relapse-free probability, so they do not coincide at
$\kappa = 0$.

```{r}
weibull_samples <- weibull_data$generate(n_replicates = n_replicates)
c(theta_T = weibull_data$treatment_effect,
  mean_estimate = mean(weibull_samples$treatment_effect_estimate, na.rm = TRUE))
```


## Generation of aggregate data with sampling approximation

### Usage

```{r}
approximate_config <- case_study_config
approximate_config$sampling_approximation <- TRUE

approximate_data <- TargetDataFactory$new()$create(
  source_data = source_data,
  case_study_config = approximate_config,
  target_sample_size_per_arm = target_sample_size_per_arm,
  control_drift = control_drift,
  treatment_drift = treatment_drift,
  summary_measure_likelihood = source_data$summary_measure_likelihood
)

samples <- approximate_data$generate(n_replicates = n_replicates)
```
```{r}
approximate_data$plot_sample(samples)
```

### Implementation details

  
It may be simpler for evaluating the OCs to sample the summary data from a
normal distribution directly, instead of sampling data using the true data
generating process. This branch keeps the simpler follow-up model in which every
patient is observed for `max_follow_up_time`: it samples the number of events in
each arm from a Poisson distribution, forms
$\operatorname{SE}(\log RR) = \sqrt{1/e_c + 1/e_t}$, and draws the summary
measure from a normal distribution. It therefore ignores the staggered accrual,
the database lock and the loss to follow-up, and is not the design described
above.

See [Data_generation_botox](../doc/Data_generation_botox.html) for details on
aggregate normal data generation.
