---
title: "Commensurate Prior"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: cerulean
    toc: yes
    toc_float:
      collapsed: true
vignette: >
  %\VignetteIndexEntry{Commensurate Prior}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{BExTE, ggplot2}
---

## Introduction

This vignette describes an implementation of the Commensurate Prior, the
location commensurate model of [Hobbs et al
(2011)](https://onlinelibrary.wiley.com/doi/10.1111/j.1541-0420.2011.01564.x)
*without* its power parameter. See the Commensurate Power Prior vignette for
the variant that carries one.

## Set working directory, import functions and configurations

List packages to load, and install them if necessary.
```{r setup,warning=TRUE, include=FALSE}
root <- dirname(dirname(getwd()))
knitr::opts_knit$set(root.dir = root, warning = FALSE, message = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

library(BExTE)
library(ggplot2)
library(cmdstanr)
```

## Mathematical description of the model

The source data inform $\theta_S$ through their own likelihood, undiscounted,
and the target parameter is tied to it by a commensurability precision $\tau$:

$$ \pi(\theta_T, \tau| \mathbf{D_S}) 
 = \int \pi(\theta_T|\theta_S, \tau) \frac{\mathcal{L}(\theta_S | \mathbf{D}_S) \pi_0(\theta_S)}{\int \mathcal{L}(\theta_S | \mathbf{D}_S) \pi_0(\theta_S) d\theta_S}d\theta_S \times p(\tau) 
$$

where $\pi_0(\theta_S)$ is an initial prior for $\theta_S$ and

$$
\theta_T|\theta_S, \tau \sim \mathcal{N}\left(\theta_S, \frac{1}{\tau}\right).
$$

This is the commensurate power prior at $\gamma \equiv 1$: the Beta prior on the
power parameter disappears, and with it the second route by which that model
adapts the amount of borrowing. Here the borrowing is governed by $p(\tau)$
alone. Small $\tau$ inflates the variance of the prior for $\theta_T$ and so
discounts the source; large $\tau$ pulls $\theta_T$ towards $\theta_S$.

## Gaussian case

With a noninformative initial prior and a Gaussian likelihood, integrating out
$\theta_S$ leaves a prior that is normal conditional on $\tau$:

$$
\theta_T \mid \mathbf{D}_S, \tau \sim \mathcal{N}\left(\hat{\theta}_S,\ \frac{1}{\tau} + \frac{\hat{\sigma}_S^2}{N_S}\right),
$$

and equation (9) of Hobbs et al (2011), evaluated at $\gamma = 1$, gives

$$ p\left(\theta_T \mid \mathbf{D}_{S}, \mathbf{D}_T, \tau, \sigma^2\right) \propto N\left(\theta_T \left\lvert\, \frac{N_S \tau \sigma^2 \hat{\theta}_S+N_T u \hat{\theta}_T}{N_S \tau \sigma^2+N_T u}\right., \frac{u \sigma^2}{N_S \tau \sigma^2+ N_T u}\right) 
$$

where $u=N_S+\hat{\sigma}_S^2 \tau$, and

$$
p\left(\tau \mid \mathbf{D}_S, \mathbf{D}_T, \sigma^2\right) \propto  N\left(\hat{\theta}_T-\hat{\theta}_S \mid 0, \frac{\sigma^2}{N_T}+\frac{1}{\tau}+\frac{\hat{\sigma}_S^2}{N_S}\right) \times \pi(\tau) .
$$

Because the posterior is an ordinary conjugate update of a normal mixture over
$\tau$, the simulation does not fit this model with Stan for every replicate:
`vectorised_replicate_inference()` discretises $p(\tau)$ on a quadrature rule
and updates the whole mixture at once. The Stan program below remains the
reference implementation, and is what `inference()` uses for a single data set.

## Implementation of the model in Stan

```{r}
case_study_config <- yaml::yaml.load_file(system.file("conf/case_studies/belimumab.yml", package = "BExTE"))


source_data <- ObservedSourceData$new(case_study_config)
summary_measure_likelihood <- case_study_config$summary_measure_likelihood
endpoint <- case_study_config$endpoint

target_data <- ObservedTargetData$new(treatment_effect_estimate = case_study_config$target$treatment_effect, treatment_effect_standard_error = case_study_config$target$standard_error, target_sample_size_per_arm = as.integer(case_study_config$target$total / 2), summary_measure_likelihood = case_study_config$summary_measure_likelihood)
```

```{r}
method_parameters <- list(
  initial_prior = "noninformative", # The posterior for the source data is derived from an uninformative prior, for consistency with the other methods.
  heterogeneity_prior = list(family = "inverse_gamma", alpha = 1 / 3, beta = 1)
  # list(family = "half_normal", std_dev = 5)
  # list(family = "cauchy", location = 0, scale = 10)
)

env = "full"
config_dir <- paste0(system.file(paste0("conf/", env), package = "BExTE"), "/")
mcmc_config <- yaml::read_yaml(paste0(config_dir, "/mcmc_config.yml"))

model <- Model$new()

model <- model$create(
  case_study_config = case_study_config,
  method = "commensurate_prior",
  method_parameters = method_parameters,
  source_data = source_data,
  mcmc_config = mcmc_config
)

print(model$stan_model_code)
```

```{r, results='hide'}
model$inference(target_data)
```

In the inference method, the data are prepared for use with Stan using the same
`prepare_data()` as the commensurate power prior - the two Stan programs share a
data block:
```{r} 
data <- model$prepare_data(target_data)
```
Then, draws are sampled from the posterior following the configuration specified in mcmc_config.

```{r}
bayesplot::mcmc_hist(model$fit$draws("target_treatment_effect"), binwidth = 0.001)
```

```{r}
posterior <- model$fit$draws("target_treatment_effect")
bayesplot::color_scheme_set("blue")
bayesplot::mcmc_dens(posterior, pars = c("target_treatment_effect"))
```

Posterior distribution of the heterogeneity parameter. There is no power
parameter to plot alongside it:
```{r}
posterior <- model$fit$draws("tau")
bayesplot::color_scheme_set("blue")
bayesplot::mcmc_dens(posterior, pars = c("tau"))
```

```{r}
library(posterior)
draws_array <- as_draws_array(model$fit)
```

```{r}
library(bayesplot)
mcmc_trace(draws_array, pars = c("tau", "target_treatment_effect"))
```

```{r}
mcmc_acf(draws_array, pars = c("tau", "target_treatment_effect"))
```

```{r}
mcmc_diagnostics <- model$fit$diagnostic_summary()

n_draws <- mcmc_config$chain_length * mcmc_config$num_chains

# MCMC Effective Sample Size
mcmc_ess <- bayesplot::neff_ratio(model$fit) * n_draws

mcmc_ess <- mcmc_ess[["target_treatment_effect"]]
print(mcmc_ess)
```

```{r}
rhat_values <- bayesplot::rhat(model$fit)
rhat <- rhat_values["target_treatment_effect"]
print(rhat)
```

```{r}
# number of divergences reported is the sum of the per chain values
n_divergences <- sum(mcmc_diagnostics$num_divergent)
print(n_divergences)
```
