---
title: "Empirical Mixture Prior (Egidi)"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: cerulean
    toc: yes
    toc_float:
      collapsed: true
vignette: >
  %\VignetteIndexEntry{Empirical Mixture Prior (Egidi)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{BExTE, ggplot2}
---

## Introduction

This vignette describes an implementation of the data-dependent mixture prior of
Egidi, Pauli and Torelli. It uses the same two components as the Robust Mixture
Prior, and differs from it in one respect only: the mixture weight is not
prespecified, it is selected from the observed target data.

See the Gaussian RMP vignette for the components themselves, which are not
changed here. Holding them fixed is what makes a comparison between the two
methods a comparison of the weight rule alone.

```{r setup, warning=TRUE, include=FALSE}
root <- dirname(dirname(getwd()))
knitr::opts_knit$set(root.dir = root, warning = FALSE, message = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

library(BExTE)
library(ggplot2)
```

## Mathematical description of the model

Egidi et al. parameterise the mixture by the weight on the *weak* component:

$$
\pi_\psi(\theta_T) = \psi\, q(\theta_T) + (1 - \psi)\, p(\theta_T),
\qquad 0 \le \psi \le 1,
$$

where $p(\theta_T) = N(\mu_p, \tau_p^2)$ is the informative, source-based
component and $q(\theta_T) = N(\mu_q, \tau_q^2)$ is the weak or unit-information
component. So $\psi = 0$ means full use of the informative prior and $\psi = 1$
means complete reliance on the weak one.

The Robust Mixture Prior in this package parameterises the *same* mixture by the
weight $w$ on the informative component, so the two conventions are related by

$$ w = 1 - \psi. $$

### The prior predictive and the conflict $p$-value

For each component $j \in \{p, q\}$, the prior-predictive distribution of the
target statistic is

$$ m_j(t) = \int f_T(t \mid \theta_T)\, j(\theta_T)\, d\theta_T. $$

With a normal target likelihood $\hat\theta_T \mid \theta_T \sim N(\theta_T,
s_T^2)$ this is available in closed form,

$$ m_p(t) = N(t \mid \mu_p,\, s_T^2 + \tau_p^2),
\qquad m_q(t) = N(t \mid \mu_q,\, s_T^2 + \tau_q^2), $$

and the predictive under the mixture is $m_\psi = \psi m_q + (1 - \psi) m_p$.

The conflict $p$-value is the probability that the prior predictive puts the
statistic somewhere at least as improbable as where it landed:

$$
P_\psi(t_{\mathrm{obs}})
= \Pr_{T \sim m_\psi}\left\{ m_\psi(T) \le m_\psi(t_{\mathrm{obs}}) \right\}.
$$

The inequality is on the density of the *complete* mixture. Averaging the two
component $p$-values is a different quantity as soon as the components have
different centres, which in this study is the ordinary case: the informative
component sits at the source estimate and the weak one at $\theta_0$.

Two cases are exact. Both ends of the weight range collapse the mixture to a
single normal, so

$$
P_0 = 2\Phi\!\left(-\frac{|t_{\mathrm{obs}} - \mu_p|}{\sqrt{s_T^2 + \tau_p^2}}\right),
\qquad
P_1 = 2\Phi\!\left(-\frac{|t_{\mathrm{obs}} - \mu_q|}{\sqrt{s_T^2 + \tau_q^2}}\right),
$$

whatever the centres are. And when the centres coincide, $\mu_p = \mu_q$, the
mixture predictive is symmetric and decreasing in $|t - \mu|$, so
$P_\psi = (1 - \psi) P_p + \psi P_q$ exactly.

Otherwise $P_\psi$ is evaluated by splitting the line into the pieces on which
$m_\psi$ is monotone and integrating between the level crossings, each of which
is then available in closed form from the component distribution functions.

### Selecting the weight

With $\alpha_{\mathrm{PC}} = 0.05$,

$$
\hat\psi = \inf\left\{ \psi \in [0, 1] : P_\psi(t_{\mathrm{obs}}) \ge \alpha_{\mathrm{PC}} \right\}.
$$

For common-centred components this inverts in closed form:

$$
\hat\psi =
\begin{cases}
0, & P_p \ge \alpha_{\mathrm{PC}},\\[4pt]
\dfrac{\alpha_{\mathrm{PC}} - P_p}{P_q - P_p}, & P_p < \alpha_{\mathrm{PC}} \le P_q,\\[10pt]
1, & P_q < \alpha_{\mathrm{PC}}.
\end{cases}
$$

In the last case the weak component cannot remove the conflict. The rule still
returns $\hat\psi = 1$, but the replicate is flagged with
`conflict_unresolved = TRUE` rather than being reported as a conflict that was
resolved.

When the centres differ, $P_\psi$ is **not** guaranteed to be monotone in $\psi$:
with the components this study uses it can fall before it rises when the observed
statistic sits near the informative centre. The infimum is therefore found by
scanning upwards for the first crossing and refining it, not by bisecting the
whole interval.

### Posterior inference

The selected prior weight and the posterior component probability are different
quantities. Writing $Z_p = m_p(t_{\mathrm{obs}})$ and $Z_q = m_q(t_{\mathrm{obs}})$,
the posterior probability of the weak component is

$$ \widetilde\psi = \frac{\hat\psi Z_q}{\hat\psi Z_q + (1 - \hat\psi) Z_p}, $$

computed on the log scale. The component posteriors are the usual conjugate
updates, and the posterior is their mixture. This is the Robust Mixture Prior's
own update, reached by handing it $w = 1 - \hat\psi$.

## Interpretation

This procedure is **empirically adaptive**. The observed target data are used
first to select the mixture weight and then again to update the posterior. That
is an intentional feature of the method, not an oversight, and it is what
distinguishes it from a Robust Mixture Prior with a prespecified weight.

Consequently $\hat\psi$ is **not** a conventional prior probability chosen before
the target data were observed, and should not be reported as one. The method is
reported separately from the Robust Mixture Prior throughout.

## The Gaussian case

```{r}
case_study_config <- list(
  name = "vignette",
  endpoint = "continuous",
  summary_measure_likelihood = "normal",
  sampling_approximation = TRUE,
  theta_0 = 0,
  null_space = "left",
  source = list(
    control = 100, treatment = 100,
    treatment_effect = 0.5, standard_error = 0.12
  )
)

source_data <- SourceData$new(case_study_config)
```

The informative component is the source posterior under a flat prior, and the
weak component is the unit-information prior the RMP uses, derived from the
observed replicate.

```{r}
informative <- list(
  mean = source_data$treatment_effect_estimate,
  sd = source_data$standard_error
)

target_sample_size_per_arm <- 60
standard_error <- 0.42
weak <- list(
  mean = case_study_config$theta_0,
  sd = sqrt(standard_error^2 * target_sample_size_per_arm)
)
```

A target estimate agreeing with the source raises no conflict, so nothing is
given to the weak component:

```{r}
agreeing <- fit_egidi_mixture(
  theta_target_hat = 0.5,
  se_target = standard_error,
  informative_component = informative,
  weak_component = weak
)
c(psi_weak = agreeing$psi_weak,
  pvalue_informative = agreeing$pvalue_informative,
  posterior_mean = agreeing$posterior_mean)
```

A conflicting one moves the weight off zero, by exactly as much as it takes to
bring the conflict $p$-value up to the threshold:

```{r}
conflicting <- fit_egidi_mixture(
  theta_target_hat = -0.75,
  se_target = standard_error,
  informative_component = informative,
  weak_component = weak
)
c(psi_weak = conflicting$psi_weak,
  pvalue_informative = conflicting$pvalue_informative,
  pvalue_selected = conflicting$pvalue_selected,
  w_informative_posterior = conflicting$w_informative_posterior)
```

### The selected weight against the observed estimate

The weight is a function of the observed data alone. Plotting it against the
observed estimate shows the shape of the rule: flat at zero while the estimate
remains predictable under the informative component, then rising once it does
not.

```{r, fig.width = 7, fig.height = 4}
estimates <- seq(-1.5, 2, length.out = 400)
selection <- egidi_select_weak_weight(
  t_obs = estimates,
  s_target = standard_error,
  mu_p = informative$mean, tau_p = informative$sd,
  mu_q = weak$mean, tau_q = weak$sd,
  alpha_pc = 0.05
)

ggplot(
  data.frame(estimate = estimates, weight = 1 - selection$psi_weak),
  aes(x = estimate, y = weight)
) +
  geom_line() +
  geom_vline(xintercept = informative$mean, linetype = "dashed") +
  labs(
    x = expression(hat(theta)[T]),
    y = expression("Informative weight  " * 1 - hat(psi))
  ) +
  ylim(0, 1) +
  theme_minimal()
```

The dashed line marks the source estimate. Note that the curve is not symmetric
about it: the weak component is centred at $\theta_0 = 0$ rather than at the
source estimate, so an estimate falling short of the source is easier for the
weak component to explain than one overshooting it by the same amount.

## The binary endpoint

For the Aprepitant case study the target statistic is the pair of responder
counts $(y_c, y_t)$, and the analysis model is the RMP's: a normal mixture on the
rate difference, each component truncated to the range the control rate leaves
it, observed through two binomial arms.

The sample space is finite, so the conflict $p$-value is computed exactly by
enumeration rather than by substituting a different statistic. For each component

$$ m_j(y_c, y_t) = \int p(y_c, y_t \mid \theta_T, p_c)\, j(\theta_T, p_c)\,
   d\theta_T\, dp_c, $$

which is tabulated over $y_c = 0, \ldots, N_T^{(c)}$ and
$y_t = 0, \ldots, N_T^{(t)}$, and

$$
P_\psi(y_c^{\mathrm{obs}}, y_t^{\mathrm{obs}})
= \sum_{\{(y_c, y_t)\, :\, m_\psi(y_c, y_t) \le m_\psi(y_c^{\mathrm{obs}}, y_t^{\mathrm{obs}})\}}
  m_\psi(y_c, y_t).
$$

```{r}
informative_table <- egidi_binomial_predictive_table(
  mu = 0.1315456, sd = 0.03505291, n_control = 52, n_treatment = 55
)
weak_table <- egidi_binomial_predictive_table(
  mu = 0, sd = 0.6, n_control = 52, n_treatment = 55
)

# The tables enumerate the whole sample space, so they sum to one.
c(informative = sum(informative_table), weak = sum(weak_table))
```

```{r}
egidi_select_weak_weight_binomial(
  table_informative = informative_table,
  table_weak = weak_table,
  y_control = 42, y_treatment = 48,
  alpha_pc = 0.05
)
```

## Configuration

The method is registered as `egidi_empirical_mixture`. It has no weight to
sweep, so it contributes a single parameter combination:

```
egidi_empirical_mixture:
  alpha_pc: 0.05
  pvalue_method: exact
  weight_grid_step: 0.001
  empirical_bayes: TRUE
```

`alpha_pc` is 0.05 in the primary analysis; 0.01 and 0.10 are the sensitivity
values.
