---
title: "KL-calibrated Normalized Power Prior"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: cerulean
    toc: yes
    toc_float:
      collapsed: true
vignette: >
  %\VignetteIndexEntry{KL-calibrated Normalized Power Prior}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{BExTE, ggplot2}
---

## Introduction

The normalised power prior of [Pawel et al
(2023)](https://doi.org/10.1002/sta4.591) discounts the source likelihood by a
power parameter $\gamma$ and places a $\operatorname{Beta}(p, q)$ prior on it.
The NPP vignette takes that prior from the configuration: a mean and a standard
deviation, chosen in advance and used for every scenario in a run.

This vignette describes `NPP_KL`, which calibrates the prior instead. The two
methods share everything downstream of it — the same joint and marginal
posteriors, the same quadrature mixture, the same summaries and effective
sample sizes. Only where $p$ and $q$ come from differs.

The motivation is that a prior fixed in advance cannot express *when* to borrow.
Whether a given source/target discrepancy is even distinguishable from noise
depends on how precise the target trial is, so a prior that ignores the target
sample size cannot say "borrow when the two studies agree, and stop borrowing at
a discrepancy I would not tolerate". The calibration states that intention
directly and solves for the shape parameters it implies.

## The criterion

Write $a$ and $b$ for the shape parameters, to keep them apart from the
significance level and the type I error. For a hypothetical target estimate $x$,
the marginal posterior of the discounting parameter is

$$
p(\gamma \mid x, D_S, a, b) \propto
\mathcal{N}\!\left(x \mid \hat\theta_S,\, s_T^2 + \frac{s_S^2}{\gamma}\right)
\operatorname{Beta}(\gamma \mid a, b), \qquad 0 < \gamma < 1,
$$

with $s_T$ the standard error the *target design* implies — not the standard
error of any particular replicate.

Two hypothetical estimates stand for the two situations the prior has to tell
apart. Under perfect compatibility the target estimate falls on the source
estimate,

$$\hat\theta_T^{(0)} = \hat\theta_S,$$

and at the maximum tolerable discrepancy it falls $d_{\mathrm{MTD}}$ away from
it, on the side of the null,

$$\hat\theta_T^{(\mathrm{MTD})} = \hat\theta_S - \texttt{benefit\_sign} \times d_{\mathrm{MTD}}.$$

Writing $p_0$ and $p_{\mathrm{MTD}}$ for the two posteriors these imply, the
calibration minimises

$$
K(a, b) = \lambda \operatorname{KL}\!\left[p_0 \,\|\, \operatorname{Beta}(c, 1)\right]
+ (1 - \lambda) \operatorname{KL}\!\left[p_{\mathrm{MTD}} \,\|\, \operatorname{Beta}(1, c)\right].
$$

$\operatorname{Beta}(c, 1)$ puts its mass near one, so the first term asks the
posterior to borrow when the studies agree; $\operatorname{Beta}(1, c)$ puts its
mass near zero, so the second asks it not to borrow at the discrepancy the user
has declared intolerable. $\lambda$ says which of the two matters more.

### Defaults

| Setting | Default | Meaning |
|---|---|---|
| `lambda_kl` | `0.5` | weight on the compatible term |
| `c_target` | `10` | shape of both reference Beta distributions |
| `d_mtd_rule` | `"source_to_null"` | $d_{\mathrm{MTD}} = k_{\mathrm{MTD}}\,\lvert\hat\theta_S - \theta_0\rvert$ |
| `d_mtd_multiplier` | `1` | the $k_{\mathrm{MTD}}$ above; `0.5` is the sensitivity setting |
| `beta_parameter_bounds` | `c(0.05, 100)` | bounds on $a$ and $b$ |

An explicit numeric `d_mtd` overrides the rule. `benefit_sign` is derived from
the case study's `null_space`: `"left"` gives $+1$, because the trial succeeds
when the effect is large, and `"right"` gives $-1$. It can also be set
explicitly.

### What the calibration does and does not read

**The hyperparameters are calibrated from two hypothetical scenarios, using the
expected target standard error.** Neither hypothetical estimate is the estimate
of any simulated replicate, and $p_0$ and $p_{\mathrm{MTD}}$ are not posteriors
from a replicate — they are the posteriors a replicate *would* have produced had
it landed exactly on the source estimate, or exactly at the maximum tolerable
discrepancy.

The calibration therefore belongs to the scenario. It is computed once, before
any replicate is generated, and every replicate of that scenario is analysed
under the same prior; only the target estimate and its standard error vary
between them. In a simulation this happens in
`Model$calibrate_for_design()`, which the drivers call once per scenario.

## Set working directory, import functions and configurations

```{r setup, warning=TRUE, include=FALSE}
root <- dirname(dirname(getwd()))
knitr::opts_knit$set(root.dir = root, warning = FALSE, message = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

library(BExTE)
library(ggplot2)
```

## Load a case study

```{r load_config}
set.seed(42)
case_study_config <- yaml::yaml.load_file(
  system.file("conf/case_studies/belimumab.yml", package = "BExTE")
)
source_data <- ObservedSourceData$new(case_study_config)
```

## Calibrate the prior

`calibrate_npp_kl()` is the calibration on its own, without a model around it.
The expected target standard error is the one the design implies.

```{r calibrate}
target_sample_size_per_arm <- as.integer(case_study_config$target$total / 2)
se_target_expected <- case_study_config$target$standard_error

calibration <- calibrate_npp_kl(
  theta_source = source_data$treatment_effect_estimate,
  se_source = source_data$standard_error,
  se_target_expected = se_target_expected,
  theta_null = case_study_config$theta_0,
  benefit_sign = benefit_sign_from_null_space(case_study_config$null_space)
)

data.frame(
  quantity = c("alpha_gamma", "beta_gamma", "prior mean", "prior sd",
               "d_MTD", "K(a, b)"),
  value = signif(c(
    calibration$alpha_gamma, calibration$beta_gamma,
    calibration$alpha_gamma / (calibration$alpha_gamma + calibration$beta_gamma),
    sqrt(calibration$alpha_gamma * calibration$beta_gamma /
           ((calibration$alpha_gamma + calibration$beta_gamma)^2 *
              (calibration$alpha_gamma + calibration$beta_gamma + 1))),
    calibration$d_mtd, calibration$objective_value
  ), 4)
)
```

The two terms pull in opposite directions, so it is worth seeing each on its
own. With all the weight on compatibility the criterion is minimised by making
the posterior of $\gamma$ equal $\operatorname{Beta}(c, 1)$, and with none of it
the answer moves towards $\operatorname{Beta}(1, c)$.

```{r lambda_sweep}
lambda_values <- seq(0, 1, by = 0.25)
sweep <- do.call(rbind, lapply(lambda_values, function(lambda) {
  fit <- calibrate_npp_kl(
    theta_source = source_data$treatment_effect_estimate,
    se_source = source_data$standard_error,
    se_target_expected = se_target_expected,
    theta_null = case_study_config$theta_0,
    benefit_sign = benefit_sign_from_null_space(case_study_config$null_space),
    lambda_kl = lambda
  )
  data.frame(
    lambda = lambda,
    alpha_gamma = fit$alpha_gamma,
    beta_gamma = fit$beta_gamma,
    prior_mean = fit$alpha_gamma / (fit$alpha_gamma + fit$beta_gamma)
  )
}))
sweep
```

```{r lambda_plot, fig.width = 6, fig.height = 4}
ggplot(sweep, aes(x = lambda, y = prior_mean)) +
  geom_line() +
  geom_point() +
  labs(
    x = expression(lambda),
    y = expression(paste("Prior mean of ", gamma)),
    title = "Where the weight on compatibility puts the prior"
  ) +
  theme_minimal()
```

## Build the model

The model is created like any other, and then calibrated to the design. Until it
is, it has no prior on the power parameter and refuses to run inference — that
is deliberate, so a model used outside a simulation driver cannot quietly
analyse under a placeholder.

```{r model}
method_parameters <- list(
  initial_prior = list("noninformative"),
  lambda_kl = list(0.5),
  c_target = list(10),
  d_mtd_rule = list("source_to_null"),
  d_mtd_multiplier = list(1)
)

model <- Model$new()$create(
  case_study_config = case_study_config,
  method = "NPP_KL",
  method_parameters = method_parameters,
  source_data = source_data
)

target_data <- TargetDataFactory$new()$create(
  source_data = source_data,
  case_study_config = case_study_config,
  target_sample_size_per_arm = target_sample_size_per_arm,
  control_drift = 0,
  treatment_drift = 0,
  summary_measure_likelihood = case_study_config$summary_measure_likelihood
)

model$calibrate_for_design(target_data)
c(p = model$p, q = model$q)
```

## Borrowing as a function of drift

Under the calibrated prior, the posterior mean of $\gamma$ falls as the target
estimate moves away from the source estimate. The prior itself does not change
along this curve: it was fixed by the design before any of these estimates was
considered.

```{r drift}
observed_target_data <- ObservedTargetData$new(
  treatment_effect_estimate = case_study_config$target$treatment_effect,
  treatment_effect_standard_error = se_target_expected,
  target_sample_size_per_arm = target_sample_size_per_arm,
  summary_measure_likelihood = case_study_config$summary_measure_likelihood
)

target_treatment_effect_values <- seq(
  source_data$treatment_effect_estimate - 2 * calibration$d_mtd,
  source_data$treatment_effect_estimate + calibration$d_mtd,
  length.out = 60
)

posterior_means <- vapply(target_treatment_effect_values, function(estimate) {
  observed_target_data$sample$treatment_effect_estimate <- estimate
  model$inference(observed_target_data)
  model$posterior_parameters$power_parameter_mean
}, numeric(1))

plot_data <- data.frame(
  drift = target_treatment_effect_values - source_data$treatment_effect_estimate,
  power_parameter_mean = posterior_means
)
```

```{r drift_plot, fig.width = 6, fig.height = 4}
ggplot(plot_data, aes(x = drift, y = power_parameter_mean)) +
  geom_line() +
  geom_vline(
    xintercept = c(0, -calibration$d_mtd),
    linetype = "dashed", colour = "grey40"
  ) +
  labs(
    x = "Drift",
    y = expression(paste("Posterior mean of ", gamma)),
    title = "Borrowing falls away as the target moves towards the null",
    subtitle = "Dashed lines: the two scenarios the prior was calibrated on"
  ) +
  theme_minimal()
```

## Reported quantities

In a simulation, every result row carries the calibration alongside the usual
posterior summaries, inside `posterior_parameters`: `alpha_gamma`, `beta_gamma`,
`prior_gamma_mean`, `prior_gamma_sd`, `d_mtd`, `se_target_expected`,
`kl_objective_value` and `calibration_converged`. They are constant across the
replicates of a scenario, which is what makes it checkable that the prior really
was fixed before the data were seen.
